Where Do Events Come From?

### 2018-08-03 paper.tex
...

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment which generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop:
from a sensor input source and back to an actuator output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{interrupt service routine,
          sleep mode,
          synchronous reactive programming}

@@

Reactive applications interact continuously and in real time with the external
world through sensors and actuators peripherals (e.g., buttons, displays,
timers, etc.).
%
These interactions are typically represented as input events flowing from the
peripherals to the application and as output events flowing from the
application to the peripherals.
%
As illustrated in Figure~\ref{fig.env}, all peripherals can be encapsulated as
a single component, \emph{the environment}, which controls the system execution
in an event loop:
the application sits idle waiting for an input;
the environment awakes the application on the occurrence of an input;
the application reacts to the input and generates back one or more outputs;
the application becomes idle and the loop restarts.

@@

The environment is typically implemented in a host language (e.g., C) and
controls the main event loop, invoking entry points in the reactive language
runtime on the occurrence of inputs and also receiving output calls, both
through a documented API.
%
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript in commands and listening for values in
subscriptions~\cite{frp.elm.ports}.

The event-based interface between the application and the environment is
arguably inevitable and also happens at the appropriate layer, since it
connects the application with the operating system resources through a
non-invasive API.
%
However, this separation exposes the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

@@

The lack of an OS opens an opportunity to explore a tighter integration between
the application and its device drivers at the language level, resulting in an
overall simplicity and tractability of the system.
%
In particular, \CEU is already amenable to a simple static analysis to detect
race conditions which we extended to also encompass ISRs.
Applications can share data buffers with drivers to avoid unnecessary copying,
with the static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts and turn off peripherals completely.
For instance, we can delimit the scope of a driver similarly to a
lexically-scoped variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in an idle states susceptible to standby.
With the help of drivers and a power manager, we can put the microcontroller to
sleep automatically at optimal sleeping modes after each input reaction.

We implemented an extensible runtime that exposes hooks for the interrupts and
the power manager, which we validated in two microcontrollers: an
\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We also implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
The applications built on top of these drivers show significant energy savings
due to automatic standby.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that toggles the state of the LED whenever a radio
           packet is received, terminating on a button press, always with the
           LED off.},
  label={lst.ceu},
]
output high/low LED;
input  high/low BUTTON;
input  Packet   RADIO_RECV;
par/or do
    var high/low v = await BUTTON (until v==low);
with
    finalize with
        emit LED(low);
    end
    loop do
        emit LED(high);
        await RADIO_RECV;
        emit LED(low);
        await RADIO_RECV;
    end
end
\end{lstlisting}

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition to run two lines of
execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED will be unconditionally
turned off~(ln~8).

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed, not only the toggling loop will be
aborted, but the \code{finalize} clause will turn the LED off unconditionally,
since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
in order to affect each other.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} suggests this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In the example, even if two packets arrive simultaneously in the environment,
the synchronous model ensures atomicity and responsiveness:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

In order to guarantee responsiveness, the synchronous model relies on the
\emph{synchronous hypothesis}, which states that reactions must be
significantly faster than the rate of inputs.
%
For this reason, \CEU (like most synchronous languages) restricts itself and
refuses unbounded loops at compile time.
This guarantees that all reactions to the environment are computed in bounded
time~\cite{ceu.lctes18}, ensuring that applications are always responsive to
incoming events.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env} to fit all
inside the same \CEU application, now with the original application represented
as the synchronous side and the environment as the asynchronous side.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers which emit input events to regular synchronous code in \CEU exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;     // connected to an LED
input  high/low PIN_02;     // connected to a button
#include "gpio.ceu"

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to virtually connect a button (pin 02) to an LED (pin 13) such that the
LED is on whenever the button is pressed and off whenever it is unpressed.

Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
with the external world and includes the asynchronous side as a driver
(ln~1--3).
It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
whenever the button changes (ln~7), changes the state of the LED with the new
value (ln~8).

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the output and
input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) and interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries in embedded
toolchains, such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically emit an input event inside its body to awake the
synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with all complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each architecture requires a mapping between the \code{async/isr} and
the actual interrupt vector table, which we provide with our system.

As we discuss in the next section, supporting ISRs in the same language and
memory space of the application allows programmers to choose between efficiency
and safety during development by providing multiple interfaces with different
levels of abstraction.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.adc},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
driver for the A/D converter.
Note that the driver (ln~1--14) and application (ln~16--27) are actually in
separate files.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do--end} scope to execute the driver (ln~21) and
use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever it goes out of scope (ln~24).
This code completely disables all A/D functionality to save energy.
%
As discussed for Listing~\ref{lst.ceu}, the finalizer would also sensibly
execute due to a \code{par/or} termination enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block and
thus cannot be inadvertently used outside it.

@@

In \CEU, when multiple trails awake in the same reaction, they execute in
lexical order, i.e., in the order they appear in the program source code.
%
In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously in accordance with the synchronous model.
Hence, for the sequence of events \code{A->B}, \code{x} becomes
\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order, the execution is still
deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).

Regardless of deterministic execution, note that an apparently innocuous change
in the order of trails may modify the behavior of programs.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
The static checks are optional (or raise warnings) and do not affect the
semantics of the language.

@@

The addition of asynchronous ISRs poses real threats concerning race conditions
in programs since interrupts occur at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a circular buffer to avoid missing data (ln~4).
The buffer is indispensable because the microcontroller cannot cope with the
USART speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signal the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

@@

Even so, we do not expect that application programmers are required to resolve
race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
the concurrency complexity inside the driver.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) with a local
buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
access to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
until the receiving the request number of bytes (ln~22).
%
On the one hand the extra copying incurs an overhead, but on the other hand it
prevents race conditions with the ISR.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.env},
]
evt_t queue[MAX];           // input queue
void main () {
  ceu_start();              // "boot reaction"
  while (1) {
    evt_t evt;
    if (ceu_input(&evt)) {  // query input queue
      ceu_sync(&evt);       // execute synchronous
    } else {
      ceu_pm_sleep();       // nothing to execute
    }
  }
}
\end{lstlisting}

Our runtime of \CEU implements the event loop proposed in
Figure~\ref{fig.async}, which alternates between synchronous, well-behaved
execution, and asynchronous ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, we want the runtime to enter in sleep mode automatically to
save energy.

Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program to spawn the ISRs and reach the first \code{await}
statements in the multiple lines of code.
%
Then, the runtime enters an infinite loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop, but they typically only
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

@@

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying the vector every time
\code{ceu\_pm\_sleep} is called.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.
           \emph{(The power manager, driver and application code are in separate files.)}},
  label={lst.pm},
  escapechar=\%,
]
// POWER MANAGER (in C)

>>>>

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop:
from a sensor input source and back to an actuator output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{
    interrupt service routines,
    reactive programming,
    sleep modes,
    standby,
    synchronous/asynchronous execution
}

@@

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
As illustrated in Figure~\ref{fig.env}, peripherals can be encapsulated as a
single component, \emph{the environment}, which connects with the application
in an event loop:
the application sits idle until the environment awakes it on the occurrence of
an input;
the application reacts to the input and generates back one or more outputs
which affect the environment;
the application becomes idle and the loop restarts.

@@

The environment is typically implemented in a host language (e.g., C) and
controls the main event loop, invoking entry points into the reactive language
runtime on the occurrence of inputs, and also receiving output calls from it,
both through a documented API.
%
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript as commands and listening for values as
subscriptions~\cite{frp.elm.ports}.

The event-based interface between the application and the environment is
arguably inevitable and also happens at the appropriate layer, since it
connects the application with the operating system resources through a
non-invasive API.
%
However, this separation reveals the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

@@

The lack of an OS opens an opportunity to explore a tighter integration between
the application and its device drivers at the language level, resulting in an
overall simplicity and tractability of the system.
%
In particular, \CEU is already amenable to a simple static analysis to detect
race conditions which we extended in this work to also encompass ISRs.
Applications can share data buffers with drivers to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts and turn off peripherals completely.
For instance, we can delimit the scope of a driver similarly to a
lexically-scoped variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, we can put the microcontroller to
sleep automatically at optimal sleeping modes after each input reaction.

We implemented an extensible runtime that exposes hooks for the interrupts and
power manager, which we validated in two microcontrollers: an
\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We also implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
The applications built on top of these drivers show significant energy savings
due to automatic standby.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that toggles the state of the LED whenever a radio
           packet is received, terminating on a button press, always with the
           LED off.},
  label={lst.ceu},
]
output high/low LED;
input  high/low BUTTON;
input  Packet   RADIO_RECV;
par/or do
    var high/low v = await BUTTON until (v==low);
with
    finalize with
        emit LED(low);
    end
    loop do
        emit LED(high);
        await RADIO_RECV;
        emit LED(low);
        await RADIO_RECV;
    end
end
\end{lstlisting}

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
of execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED will be unconditionally
turned off~(ln~8).

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
in order to affect each other.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} suggests this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures atomicity and responsiveness in the
application:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

In order to guarantee responsiveness, the synchronous model relies on the
\emph{synchronous hypothesis}, which states that reactions must be
significantly faster than the rate of inputs.
%
For this reason, \CEU (like most synchronous languages) refuses unbounded loops
at compile time.
This restriction guarantees that all reactions to the environment are computed
in bounded time~\cite{ceu.lctes18}, ensuring that applications are always
responsive to upcoming events.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
original application represented as the synchronous side and the environment as
the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;  // connected to an LED
input  high/low PIN_02;  // connected to a button
#include "gpio.ceu"      // GPIO driver

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to connect a button (pin 02) to an LED (pin 13) via software such that
the LED is on whenever the button is pressed and off whenever it is unpressed.

Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
with the external world and includes the asynchronous side as a driver
(ln~1--3).
It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
whenever the button changes (ln~7), toggles the state of the LED with the new
value (ln~8).

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains, such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically emit an input event inside its body to awake the
synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with the complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each supported architecture requires a mapping between the
\code{async/isr} and the actual interrupt vector table (which we provide
for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The A/D driver (ln~1--14) is included in the application inside a
           lexically-scoped block (ln~20--24), which turns off all driver
           functionalities automatically on termination.},
  label={lst.adc},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
driver for the A/D converter.
Note that the driver (ln~1--14) and application (ln~16--27) are actually in
separate files.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The finalize completely disables all A/D functionality to save energy.
%
As discussed in the previous section, the finalizer would also execute if
aborted from a \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block and
thus cannot be inadvertently used outside it.

@@

In \CEU, when multiple trails awake in the same reaction, they execute in
lexical order, i.e., in the order they appear in the source code.
%
In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously in accordance with the synchronous model.
%Hence, for the sequence of events \code{A->B}, \code{x} becomes
%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
%\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order, the execution is still
deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
Even so, \CEU performs (optional) concurrency checks at compile time to detect
conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
%The static checks are optional (or raise warnings) and do not affect the
%semantics of the language.

@@

The addition of asynchronous ISRs poses real threats concerning race conditions
since they interrupt programs at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a circular buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller may cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

@@

However, we do not expect that application programmers should be required to
resolve race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
the concurrency complexity inside the driver.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) passing a
local buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
access to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
up to the requested number of bytes (ln~22).
%
On the one hand the extra copying incurs a memory and runtime overhead, but on
the other hand it prevents race conditions with the ISR.

Support for ISRs in the same language and memory space of the application
allows programmers to choose between efficiency and safety during development
by providing multiple interfaces with different levels of abstraction.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Overall runtime architecture of \CEU with an input queue (ln~1),
           event loop (ln~4--11), synchronous execution (ln~7), and standby
           mode (ln~9).},
  label={lst.env},
]
evt_t queue[MAX];           // input queue
void main () {
  ceu_start();              // "boot reaction"
  while (1) {
    evt_t evt;
    if (ceu_input(&evt)) {  // query input queue
      ceu_sync(&evt);       // execute synchronous
    } else {
      ceu_pm_sleep();       // nothing to execute
    }
  }
}
\end{lstlisting}

Our runtime implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, it is possible for the runtime to enter in sleep mode
automatically to save energy.

Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple lines of code.
%
Then, the runtime enters an infinite loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop, but they typically only
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

@@

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying this vector every time
\code{ceu\_pm\_sleep} is called.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Interaction between the power manager, USART driver, and application
           (each code is actually in a separate file).},
  label={lst.pm},
  escapechar=\%,
]
// POWER MANAGER (in C)
